HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-27 scnn.tex
small edit

We make full sky N-body simulations for two parameter sets for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ \todo{(MORE PARAMS HERE - CHECK WITH RAPHAEL)}.
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel of radius $3$ arcmins. The resulting PSD are displayed in \figref{psd_sigma3}. In the pre-processing step, we also remove the mean of each map and downsample them to a resolution of of $Nside=1024$.

@@

We compare the performance of the spherical CNN against two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities (PSD) and b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified using a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better results
while having significanly worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailed in Appendix~\ref{sec:dataset_augmentation}. We tuned the regularization parameter by cross-validating with the validation set.

@@

\subsection{Filter visualization}
One common visualization for CNN is to observe the shape of the learned filters.
Since our construction leads to almost (up to the sampling irregularities)
spherical filters, we plot in \figref{learned_filter} both the section (on the
sphere) and a gnomonic projection on a plane of a random selection of learned
filters. The filters are selected from the last layer of the network
corresponding the experiment with order $o=2$ for a relative noise level of
$2$. Nevertheless, all networks presented similar gnomonic and section pattern, at least from a
human perspective. The gnomonic projection clearly exibits the sampling issue
of HEALPix.
For a detail, on how the convolution kernels are plotted, please refer to Appendix~\ref{sec:filter_visualization}.
While it is usually difficult to interpret the shape of the filters in the context of the type of data, we can notice that they often have a ``peak'' like structures.
An example of filter interpretation was demonstrated in \citep{Ribli2018learningfrom}.
% \todo{Tomek: can you say something about the filters.}
% \TK{Done }

@@

\section{Augmentation of the dataset for the SVM classifier}
\label{sec:dataset_augmentation} The original number of samples $480$, $1920$,
$7680$ respectively for order $o=1,2,4$ is considered rather low for usual
deep learning technique, especially when one desire to build a noise robust
classifier. As a result, we use a classical dataset augmentation technique.
Before entering the network for training each sample is added a random
realization of Gaussian noise. The effect is twofold. First, the network never
ends up seing the same sample and is less likely to overfit the training set.
Second the trained network will be robust to Gaussian noise.

When comparing the network, we want to ensure that the SVM classifier has
access to the same amount of data that the network, i.e, potentially an
infinite number of samples. Hence, we fit different linear SVM classifier using
various training set size until we observe experimentally that
increasing the amount of data does not improve the classifier. The validation
set remain the same and is used for cross-validation to tune the L2
penalization parameter of SVM. Eventually, we evaluate the final classifier on
the testing set. An example of the evolution of the training and valdication
error is shown in \figref{hist_error_evolution}.

>>>>

We make full sky N-body simulations for two parameter sets for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ \todo{(MORE PARAMS HERE - CHECK WITH RAPHAEL)}.
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel of radius $3$ arcmins. The resulting PSD are displayed in \figref{psd_sigma3}. In the pre-processing step, we also remove the mean of each map and downsample them to a resolution of of $N_{side}=1024$.

@@

We compare the performance of the spherical CNN against two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities (PSD) and b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified using a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better results
while having significanly worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailed in \ref{sec:dataset_augmentation}. We tuned the regularization parameter by cross-validating with the validation set.

@@

\subsection{Filter visualization}
One common visualization for CNN is to observe the shape of the learned filters.
Since our construction leads to almost (up to the sampling irregularities)
spherical filters, we plot in \figref{learned_filter} both the section (on the
sphere) and a gnomonic projection on a plane of a random selection of learned
filters. The filters are selected from the last layer of the network
corresponding the experiment with order $o=2$ for a relative noise level of
$2$. Nevertheless, all networks presented similar gnomonic and section pattern, at least from a
human perspective. The gnomonic projection clearly exibits the sampling issue
of HEALPix.
For a detail, on how the convolution kernels are plotted, please refer to \ref{sec:filter_visualization}.
While it is usually difficult to interpret the shape of the filters in the context of the type of data, we can notice that they often have a ``peak'' like structures.
An example of filter interpretation was demonstrated in \citep{Ribli2018learningfrom}.
% \todo{Tomek: can you say something about the filters.}
% \TK{Done }

@@

\section{Augmentation of the dataset for the SVM classifier}
\label{sec:dataset_augmentation} The original number of samples $480$, $1920$,
$7680$ respectively for order $o=1,2,4$ is considered small to train deep a architecture, 
especially when one desire to build a noise robust
classifier. As a result, we use a classical dataset augmentation technique.
Before entering the network for training each sample is added a random
realization of Gaussian noise. The effect is twofold. First, the network never
ends up seing the same sample and is less likely to overfit the training set.
Second the trained network will be robust to Gaussian noise.

When comparing the network to the baselines, we want to ensure that the SVM classifier has
access to the same amount of data that the network, i.e, potentially an
infinite number of samples. Hence, we fit different linear SVM classifiers using
various training set size until we observe experimentally that
increasing the amount of data does not improve the accuracy. The validation
set remain the same and is used for cross-validation to tune the L2
penalization parameter of SVM. Eventually, we evaluate the final classifier on
the testing set. An example of the evolution of the training and valdication
error is shown in \figref{hist_error_evolution}.
